Efficient simulation of individual-based population models

### 2024-10-11 article.qmd
computo second review

---
title: "Efficient simulation of individual-based population models"
subtitle: "The `R` package `IBMPopSim`"
author:
  - name: Daphné Giorgi 
    corresponding: true
    email: daphne.giorgi@sorbonne-universite.fr
    url: https://perso.lpsm.paris/~giorgi/
    affiliations:
      - name: Sorbonne Université, CNRS
        department: Laboratoire de Probabilités, Statistique et Modélisation
        url: https://www.lpsm.paris/
  - name: Sarah Kaakai
    corresponding: true
    email: sarah.kaakai@univ-lemans.fr
    url: https://perso.univ-lemans.fr/~skaakai/
    affiliations:
      - name: Le Mans Université
        department: Laboratoire Manceau de Mathématiques
        url: https://lmm.univ-lemans.fr/fr/
      - name: Ecole Polytechnique, CNRS
        department: Centre de Mathématiques Appliquées
        url: https://cmap.ip-paris.fr/ 
  - name: Vincent Lemaire 
    corresponding: true
    email: vincent.lemaire@sorbonne-universite.fr
    url: https://perso.lpsm.paris/~vlemaire/
    orcid: 0000-0002-0433-7722
    affiliations:
      - name: Sorbonne Université, CNRS
        department: Laboratoire de Probabilités, Statistique et Modélisation
        url: https://www.lpsm.paris/
date: last-modified
date-modified: last-modified
description: |
    This document provides a full description of the Stochastic Individual-Based Models (IBMs) that can be implemented in the IBMPopSim package. A unified mathematical and simulation framework is given, with a detailed description of the simulation algorithm. Examples of applications for the package are also provided, showing the performance and flexibility of IBMPopSim. 
abstract: >+
    The `R` Package `IBMPopSim` aims to simulate the random evolution of heterogeneous populations using stochastic Individual-Based Models (IBMs). The package enables users to simulate population evolution, in which individuals are characterized by their age and some characteristics, and the population is modified by different types of events, including births/arrivals, death/exit events, or changes of characteristics. The frequency at which an event can occur to an individual can depend on their age and characteristics, but also on the characteristics of other individuals (interactions). Such models have a wide range of applications in fields including  actuarial science, biology, ecology or epidemiology. `IBMPopSim` overcomes the limitations of time-consuming IBMs simulations by implementing new efficient algorithms  based on thinning methods, which are compiled using the `Rcpp` package while providing a user-friendly interface.
keywords: [Individual-based models, stochastic simulation, population dynamics, Poisson measures, thinning method, actuarial science, insurance portfolio simulation]
citation:
  type: article-journal
  container-title: "Computo"
  doi: "xxxx"
  url: https://computo.sfds.asso.fr/template-computo-quarto
  publisher: "Société Française de Statistique"
  issn: "2824-7795"
bibliography: references.bib
github-user: VincentLemaire
repo: "ibmpopsim_computo"
draft: true # set to false once the build is running
published: false # will be set to true once accepted
format:
    computo-html: default
    computo-pdf: default
crossref:
    cnj-title: Assumption
    cnj-prefix: Assumption 
execute:
    eval: true
    echo: true
    warning: false
---

@@

As mentioned in Section @sec-overview a population is a collection of individuals whose evolution defines the population process.

@@

- At time $T$, the population $Z_T$ contains all individuals who lived in the population before $T$, including dead/exited individuals. If there are no swap events, or entries,the population state $Z_t$ for any time $t\leq T$ can be obtained from $Z_T$. Indeed, if $Z_T = \sum_{k=1}^{N_T}  \delta_{I_k}$, then the population at time $t\leq T$ is simply composed of the individuals born before $t$: $$
    Z_t = \sum_{k=1}^{N_T} \mathbf{1}_{\{\tau^b_k \leq t \}} \delta_{I_k}.
$$
- In the presence of entries (open population), a characteristic $x$ can track the individuals' entry dates. Then, the previous equation can be easily modified in order to obtain the population $Z_t$ at time $t\leq T$ from $Z_T$.

@@

- When the event intensity does not depend on the population state, 
$$
    \big(\Lambda^e_t (Z_t)\big)_{t\in [0,T]} = \big(\mu^e(t)\big)_{t \in [0,T]},
$$ {#eq-poisson-intensity}
with $\mu^e$ a deterministic function, the events of type $e$ occur at the jump times of an inhomogeneous Poisson process of intensity function $(\mu^e(t))_{t \in [0,T]}$. When such an event occurs, the individual to whom the event happens to is drawn uniformly among alive individuals in the population.
In a given model, the set of events $e\in E$ with Poisson intensities will be denoted by $\mathcal{P}$.
- Otherwise, we assume that the global intensity $\Lambda^e_t(Z_t)$ at which the events of type $e$ occur in the population can be written as the sum of individual intensities $\lambda^e_t(I,Z_t)$:
$$
\begin{aligned}
    &\Lambda^e_t (Z_t) = \sum_{k=1}^{N_t} \lambda^e_t ( I_k,Z_t),  \\
    &  \text{with } \mathbb P\big( \text{event } e \text { occurs to an individual } I \in (t,t+\mathrm{d}t] | \mathcal{F}_t \big) \simeq  \lambda^e_t (I,Z_t)\mathrm{d}t.
\end{aligned}
$$ {#eq-global-intensity}

@@

1. *No interactions:* The intensity function $\lambda^e$ does not depend on the population composition. The intensity at which the event of type $e$ occur to an individual $I$ only depends on its date of birth and characteristics: 
$$
    \lambda^e_t (I,Z_t) = \lambda^e(t, I),
$$ {#eq-intensity-no-interaction}
where $\lambda^e: \mathbb{R}_+ \times \mathcal{I}\to {\mathbb{R}}^+$ is a deterministic function. In a given model, we denote by $\mathcal{E}$ the set of event types with individual intensity @eq-intensity-no-interaction.

@@

2. In the presence of competition between individuals, the death intensity of an individual $I$ also depend on other individuals $J$ in the population. For example, if $I=(\tau^b,\tau^d, x)$, with its size $x$, then we have: 
$$
    W^d(t,I,J) = (x_J - x)^+ \mathbf{1}_{\{\tau^d_J > t\}}, \quad \forall \; J=(\tau^b_J,\tau^d_J , x_J).
$$ {#eq-ex-interaction}
This can be interpreted as follows: if the individual $I$ meets randomly an individual $J$ alive at time $t$, and of bigger size $x_J > x$, then he can die at the intensity $x_J-x$. If $J$ is smaller than $I$, then he cannot kill $I$. The bigger is the size $x$ of $I$, the lower is his death intensity $\lambda^d_t(I,Z_t)$ defined by 
$$
    \lambda^d_t(I,Z_t) = \sum_{\genfrac{}{}{0pt}{2}{J\in Z_t,}{x_J > x}} (x_J -x)\mathbf{1}_{\{\tau^d_J > t\}}.
$$

@@

### Events intensity bounds

@@

::: {#cnj-intensity-poisson} 
For all events $e \in \mathcal{P}$ with Poisson intensity @eq-poisson-intensity, the intensity is assumed to be bounded on $[0,T]$:
$$
    \forall t \in [0,T], \quad \Lambda^e_t(Z_t) = \mu^e(t) \leq \bar \mu^e.
$$
:::

@@

::: {#cnj-intensity1}
For all event types $e \in \mathcal{E}$, the associated individual event intensity $\lambda^e$ with no interactions, _i.e._ $\lambda^e$ verifies @eq-intensity-no-interaction, is assumed to be uniformly bounded:
$$
    \lambda^e(t, I) \leq \bar \lambda^e, \quad \forall \;  t\in [0, T],  \;   I \in \mathcal{I}.
$$
In particular, 
$$
    \forall t \in [0,T], \quad \Lambda^e_t (Z_t) = \sum_{k=1}^{N_t} \lambda^e(t, I) \leq \bar \lambda^e  N_t .
$$ {#eq-def-bar-Lambda}
:::

@@

Since the seminal paper of @FouMel04, it has been shown in many examples that a stochastic IBM dynamics can be defined rigorously as the unique solution of an SDE driven by Poisson measures, under reasonable non explosion conditions. In the following, we introduce a unified framework for the pathwise representation of the class of stochastic IBMs introduced above. 
Some recalls on Poisson random measures are presented in the Appendix @sec-recall-poisson, and for more details on these representations of particular examples, we refer to the abundant literature on the subject.

In the following we consider an individual-based stochastic population $(Z_t)_{t\in [0,T]}$, keeping the notations introduced in @sec-events and @sec-event-intensity for the events and their intensities.
In particular, the set of events types that define the population evolution is denoted by $\mathcal{P} \cup \mathcal{E} \cup \mathcal{E}_W \subset E$, with $\mathcal{P}$ the set of events types with Poisson intensity verifying @cnj-intensity-poisson, $\mathcal{E}$ the set of events types with individual intensity and no interaction, verifying @cnj-intensity1 and finally $\mathcal{E}_W$ the set of event types with interactions, verifying @cnj-intensity2.

### Non-explosion criterion

First, one has to ensure that the number of events occurring in the population will not explode in finite time, leading to an infinite simulation time. @cnj-intensity1 and @cnj-intensity2 are not sufficient to guarantee the non explosion of the event number, due to the potential explosion of the population size in the presence of interactions. An example is the case when only birth events occur, with an intensity $\Lambda^b_t(Z_t) = C_b (N_t^a)^2$ ($W^b(t, I,J) =C_b$). Then, the number of alive individuals $(N_t^a)_{t\geq 0}$ is a well-known pure birth process of intensity function $g(n) = C_b n^2$ (intensity of moving from state $n$ to $n+1$). This process explodes in finite time, since $g$ does not verify the necessary and sufficient non explosion criterion for pure birth Markov processes: $\sum_{n=1}^\infty \frac{1}{g(n)} = \infty$ (see e.g. Theorem 2.2 in [@BanMel15]). There is thus an explosion in finite time of birth events.

@@

@cnj-non-explosion yields that the global intensity $\Lambda_t^e(\cdot)$ of event $e$ is bounded by a function $g^e$ only depending on the population size:
$$
    \Lambda_t^e (Z) \leq g^e(n) := nf^e(n), \quad \text{with }\sum_{n=1}^{\infty} \frac{1}{g^e(n)} = \infty.
$$
If $e\in \mathcal{P}$ has a Poisson intensity, then $\Lambda_t^e(Z) =\mu^e_t$ always verifies the previous equation with $g^e(n) = \bar \mu^e$.

@@

The proof of @lem-non-explosion, detailed in Appendix @sec-prf-lem-non-explosion is more technical and rely on pathwise comparison result, generalizing those obtained in [@KaaElK20]. An alternative pathwise representation of the population process, inspired by the randomized @alg-randomized is given as well in @prp-randomized.

@@

We now present the main algorithm for simulating the evolution of an IBM over $[0,T]$.The algorithm implemented in `IBMPopSim` allows the exact simulation of @eq-sde-eq-Z, based on an acceptance/reject algorithm for simulating random times called *thinning*. The exact simulation of event times with this acceptance/reject procedure is closely related to the simulations of inhomogeneous Poisson processes by the so-called thinning algorithm, often attributed to @LewShe79. 
The simulation methods for inhomogeneous Poisson processes can be adapted to IBMs, and we introduce in this section a general algorithm extending those by @FouMel04 (see also @FerTra09, @Ben10).

The algorithm is based on exponential "candidate" event times, chosen with a (constant) intensity which must be greater than the global event intensity $\Lambda_t(Z_t)$ (@eq-global-intensity). Starting from time $t$, once a candidate event time $t + \bar T_\ell$ has been proposed, a candidate event type $e$ (birth, death,\...) is chosen with a probability $p^e$ depending on the event intensity bounds $\bar \mu^e$, $\bar \lambda^e$ and $\bar W^e$, as defined in @cnj-intensity1 and @cnj-intensity2. 
An individual $I$ is then drawn from the population. 
Finally, it remains to accept or reject the candidate event with a probability $q^e(t,I,Z_t)$ depending on the true event intensity. 
If the candidate event time is accepted, then the event $e$ occurs at time $t + \bar T_\ell$ to the individual $I$. The main idea of the algorithm implemented can be summarized as follows:

@@

Let $(N^j)_{j \in \mathcal{J}}$ be a (inhomogeneous) multivariate Poisson process indexed by a finite set $\mathcal{J}$, such that $\forall j \in \mathcal{J}$, the intensity $(\lambda_j(t))_{t\in [0,T]}$ of $N_j$ is bounded on $[0,T]$:
$$
    \sup_{t\in[0,T]} \lambda_j(t) \le \bar \lambda_j, \text{ and let }
    \bar \Lambda = \sum_{j \in \mathcal{J}} \bar \lambda_j.
$$ 
Recall that such multivariate counting process can be rewritten as a Poisson random measure $N= \sum_{k\geq 1} \delta_{(T_k,J_k)}$ on ${\mathbb{R}}^+\times \mathcal{J}$ (see e.g. Sec. 2 of Chapter 6 in [@Cin11]), where $T_k$ is the $k$th jump time of $\sum_{j\in \mathcal{J}} N^j$ and $J_k$ corresponds to the component of the the vector which jumps. In particular, $N^j_t = N([0,t]\times \{j\})$.

@@

\begin{algorithm}
\caption{Thinning algorithm for multivariate inhomogeneous Poisson processes.}
\begin{algorithmic}
    \State \textbf{Input:} Functions $\lambda_j: [0,T] \to [0,\bar \lambda]$ and $\bar \lambda_j$, $\bar \Lambda = \sum_{j \in \mathcal{J}} \bar \lambda_j$
    \State \textbf{Output:} Points $(T_k, J_k)$ of Poisson measure $N$ on $[0,T] \times \mathcal{J}$
    \State Initialization $T_0 \longleftarrow 0$, $\bar T_0 \longleftarrow 0$
    \While{$T_k < T$}
        \Repeat
            \State increment iterative variable $\ell \longleftarrow \ell+1$
            \State compute next proposed time $\bar T_{\ell} \longleftarrow \bar T_{\ell-1} + S_\ell$  with $S_\ell \sim \mathcal{E}(\bar \Lambda)$ 
            \State draw $\bar J_\ell \sim \mathcal{U}\big\{ \bar \lambda_j / \bar \Lambda, j \in \mathcal{J} \big\}$ i.e. $\mathbb{P}(\bar J_\ell = j) = \bar \lambda_j / \bar \Lambda$
            \State draw $\bar \Theta_\ell \sim \mathcal{U}([0,\bar \lambda_{\bar J_\ell}])$
        \Until{accepted event $\bar \Theta_\ell \le \lambda_{\bar J_\ell}(\bar T_{\ell})$} 
        \State record $(T_k, J_k) \longleftarrow (\bar T_{\ell}, \bar J_\ell)$ as accepted point
    \EndWhile
\end{algorithmic}
\end{algorithm}
```

@@

### First event simulation

@@

Since the population composition is constant before the first event time, each counting process $N^{j}$ coincides on $[0,T_1[$ with an inhomogeneous Poisson process, of intensity $\lambda_t^e (I_k,Z_0)$. Thus (conditionally to $Z_0$), $T_1$ is also the first jump time of an inhomogeneous multivariate Poisson process $N^0 = (N^{0,j})_{j \in \mathcal{J}_0}$ of intensity function $(\lambda_j)_{j\in \mathcal J_0}$, defined for all $j = (k,e) \in \mathcal{J}_0$ by:
$$
    \lambda_j(t) = \lambda^e_t(I_k,Z_0) \le \bar \lambda^e_0 \quad \text{with} \quad \bar \lambda^e_0 = \bar \lambda^e \mathbf{1}_{e \in \mathcal{E}} + \bar W^e N_0 \mathbf{1}_{e \in \mathcal{E}_W},
$$
by @cnj-intensity1 and @cnj-intensity2. In particular, the jump times of $N^0$ occur at the intensity
$$
    \Lambda(t) =\sum_{j \in \mathcal{J}_0} \lambda_j(t)  =\sum_{e \in \mathcal{E} \cup \mathcal{E}_W} \sum_{k=1}^{N_0}  \lambda^e_t(I_k,Z_0) \leq \bar \Lambda(N_0)=N_0 \sum_{e \in \mathcal{E} \cup \mathcal{E}_W} \bar \lambda^e_0.
$$
By @prp-thinning-vector, $N^0$ can be obtained by thinning of the marked Poisson measure $\bar Q^0 = \sum_{\ell \geq 1} \delta_{(\bar T_\ell , (\bar{K}_\ell, \bar E_\ell), \bar \Theta_\ell)}$ on ${\mathbb{R}}^+\times\mathcal{J}_0 \times {\mathbb{R}}^+$, with:

@@

::: {.proof}
The only difference between @alg-interaction and @alg-randomized is in the acceptance/rejection step of proposed events, in the presence of interactions.
In @alg-randomized , a proposed event $(\bar T_\ell, \bar E_\ell, \bar K_\ell)$, with $\bar E_l \in \mathcal{E}_W$ an event with interaction, is accepted as a true event in the population if
$$
    \bar \Theta_\ell \le W^{\bar E_\ell}(\bar T_\ell, I_{\bar K_\ell}, I_{\bar J_\ell}), \text{ with } (\bar \Theta_\ell, \bar J_\ell) \sim  \mathcal{U}\big([0, \bar W^{\bar E_\ell}] \times \{1, \dots, N_{T_k}\} \big).
$$
By @eq-randomized-event, the probability of accepting this event is the same than in @alg-interaction, which achieves the proof.
:::

@@

The most important step of the model creation is the events creation. The call to the function creating an event is of form
```{R}
#| eval: false 
mk_event_CLASS(type="TYPE", name="NAME", ...)
```
where `CLASS` is replaced by the class of the event intensity, described in @sec-event-intensity , and `type` corresponds to the event type, described in Section @sec-events.
@tbl-intensity-classes and @tbl-event-types summarize the different possible choices for intensity classes and types of event. The optional argument `name` gives a name to the event. If not specified, the name of the event is its type, for instance `death`. However, a name must be specified if the model is composed of several events with the same type (for instance when there are multiple death events corresponding to different causes of death). The other arguments depend on the intensity class and on the event type.

@@

The proof @thm-EqZrandomized  follows the same steps than the proof of
@thm-eq-Z.

>>>>

---
title: "Efficient simulation of individual-based population models"
subtitle: "The `R` package `IBMPopSim`"
author:
  - name: Daphné Giorgi 
    corresponding: true
    email: daphne.giorgi@sorbonne-universite.fr
    url: https://perso.lpsm.paris/~giorgi/
    affiliations:
      - name: Sorbonne Université, CNRS
        department: Laboratoire de Probabilités, Statistique et Modélisation
        url: https://www.lpsm.paris/
  - name: Sarah Kaakai
    corresponding: true
    email: sarah.kaakai@univ-lemans.fr
    url: https://perso.univ-lemans.fr/~skaakai/
    affiliations:
      - name: Le Mans Université
        department: Laboratoire Manceau de Mathématiques
        url: https://lmm.univ-lemans.fr/fr/
      - name: Ecole Polytechnique, CNRS
        department: Centre de Mathématiques Appliquées
        url: https://cmap.ip-paris.fr/ 
  - name: Vincent Lemaire 
    corresponding: true
    email: vincent.lemaire@sorbonne-universite.fr
    url: https://perso.lpsm.paris/~vlemaire/
    orcid: 0000-0002-0433-7722
    affiliations:
      - name: Sorbonne Université, CNRS
        department: Laboratoire de Probabilités, Statistique et Modélisation
        url: https://www.lpsm.paris/
date: last-modified
date-modified: last-modified
description: |
    This document provides a full description of the Stochastic Individual-Based Models (IBMs) that can be implemented in the IBMPopSim package. A unified mathematical and simulation framework is given, with a detailed description of the simulation algorithm. Examples of applications for the package are also provided, showing the performance and flexibility of IBMPopSim. 
abstract: >+
    The `R` Package `IBMPopSim` facilitates the simulation of the random evolution of heterogeneous populations using stochastic Individual-Based Models (IBMs). The package enables users to simulate population evolution, in which individuals are characterized by their age and some characteristics, and the population is modified by different types of events, including births/arrivals, death/exit events, or changes of characteristics. The frequency at which an event can occur to an individual can depend on their age and characteristics, but also on the characteristics of other individuals (interactions). Such models have a wide range of applications in fields including  actuarial science, biology, ecology or epidemiology. `IBMPopSim` overcomes the limitations of time-consuming IBMs simulations by implementing new efficient algorithms  based on thinning methods, which are compiled using the `Rcpp` package while providing a user-friendly interface.
keywords: [Individual-based models, stochastic simulation, population dynamics, Poisson measures, thinning method, actuarial science, insurance portfolio simulation]
citation:
  type: article-journal
  container-title: "Computo"
  doi: "xxxx"
  url: https://computo.sfds.asso.fr/template-computo-quarto
  publisher: "Société Française de Statistique"
  issn: "2824-7795"
bibliography: references.bib
github-user: VincentLemaire
repo: "ibmpopsim_computo"
draft: true # set to false once the build is running
published: false # will be set to true once accepted
format:
    computo-html: default
    computo-pdf: default
crossref:
    cnj-title: Assumption
    cnj-prefix: Assumption 
execute:
    eval: true
    echo: true
    warning: false
---

@@

As mentioned in @sec-overview a population is a collection of individuals whose evolution defines the population process.

@@

- At time $T$, the population $Z_T$ contains all individuals who lived in the population before $T$, including dead/exited individuals. If there are no swap events, or entries, the population state $Z_t$ for any time $t\leq T$ can be obtained from $Z_T$. Indeed, if $Z_T = \sum_{k=1}^{N_T}  \delta_{I_k}$, then the population at time $t\leq T$ is simply composed of the individuals born before $t$: $$
    Z_t = \sum_{k=1}^{N_T} \mathbf{1}_{\{\tau^b_k \leq t \}} \delta_{I_k}.
$$
- In the presence of entries (open population), a characteristic $x$ can track the individuals' entry dates. Then, the previous equation can be easily modified in order to obtain the population $Z_t$ at time $t\leq T$ from $Z_T$.

@@

- When the event intensity does not depend on the population state, 
$$
    \big(\Lambda^e_t (Z_t)\big)_{t\in [0,T]} = \big(\mu^e(t)\big)_{t \in [0,T]},
$$ {#eq-poisson-intensity}
with $\mu^e$ a deterministic function, the events of type $e$ occur at the jump times of an inhomogeneous Poisson process of intensity function $(\mu^e(t))_{t \in [0,T]}$. 
This is particularly useful when entry events occur with intensities influenced by environmental processes and/or exhibit seasonal variations.
When such an event occurs, the individual to whom the event happens is drawn uniformly from the living individuals in the population.
In a given model, the set of events $e\in E$ with Poisson intensities will be denoted by $\mathcal{P}$.
- Otherwise, we assume that the global intensity $\Lambda^e_t(Z_t)$ at which the events of type $e$ occur in the population can be written as the sum of individual intensities $\lambda^e_t(I,Z_t)$:
$$
\begin{aligned}
    &\Lambda^e_t (Z_t) = \sum_{k=1}^{N_t} \lambda^e_t ( I_k,Z_t),  \\
    &  \text{with } \mathbb P\big( \text{event } e \text { occurs to an individual } I \in (t,t+\mathrm{d}t] | \mathcal{F}_t \big) \simeq  \lambda^e_t (I,Z_t)\mathrm{d}t.
\end{aligned}
$$ {#eq-global-intensity}

@@

1. *No interactions:* The intensity function $\lambda^e$ does not depend on the population composition. The intensity at which an event of type $e$ occurs to an individual $I$ only depends on its date of birth and characteristics: 
$$
    \lambda^e_t (I,Z_t) = \lambda^e(t, I),
$$ {#eq-intensity-no-interaction}
where $\lambda^e: \mathbb{R}_+ \times \mathcal{I}\to {\mathbb{R}}^+$ is a deterministic function. In a given model, we denote by $\mathcal{E}$ the set of event types with individual intensity @eq-intensity-no-interaction.

@@

2. In the presence of competition between individuals, the death intensity of an individual $I$ also depends on other individuals $J$ in the population. For example, if $I=(\tau^b,\tau^d, x)$, with its size $x$, then we have: 
$$
    W^d(t,I,J) = (x_J - x)^+ \mathbf{1}_{\{\tau^d_J > t\}}, \quad \forall \; J=(\tau^b_J,\tau^d_J , x_J).
$$ {#eq-ex-interaction}
This can be interpreted as follows: if the individual $I$ meets randomly an individual $J$ alive at time $t$, and of bigger size $x_J > x$, then he can die at the intensity $x_J-x$. If $J$ is smaller than $I$, then it cannot kill $I$. The bigger is the size $x$ of $I$, the lower is its death intensity $\lambda^d_t(I,Z_t)$ defined by 
$$
    \lambda^d_t(I,Z_t) = \sum_{\genfrac{}{}{0pt}{2}{J\in Z_t,}{x_J > x}} (x_J -x)\mathbf{1}_{\{\tau^d_J > t\}}.
$$

@@

### Events intensity bounds {.unnumbered}

@@

::: {#cnj-intensity-poisson} 
For all events $e \in \mathcal{P}$ with Poisson intensity (@eq-poisson-intensity), the intensity is assumed to be bounded on $[0,T]$:
$$
    \forall t \in [0,T], \quad \Lambda^e_t(Z_t) = \mu^e(t) \leq \bar \mu^e.
$$
:::

@@

::: {#cnj-intensity1}
For all event types $e \in \mathcal{E}$, the associated individual event intensity $\lambda^e$ with no interactions (@eq-intensity-no-interaction) is assumed to be uniformly bounded:
$$
    \lambda^e(t, I) \leq \bar \lambda^e, \quad \forall \;  t\in [0, T],  \;   I \in \mathcal{I}.
$$
In particular, 
$$
    \forall t \in [0,T], \quad \Lambda^e_t (Z_t) = \sum_{k=1}^{N_t} \lambda^e(t, I) \leq \bar \lambda^e  N_t .
$$ {#eq-def-bar-Lambda}
:::

@@

Since the seminal paper of @FouMel04, it has been shown in many examples that a stochastic IBM dynamics can be defined rigorously as the unique solution of an SDE driven by Poisson measures, under reasonable non explosion conditions. In the following, we introduce a unified framework for the pathwise representation of the class of stochastic IBMs introduced above. 
Some recalls on Poisson random measures are presented in the Appendix @sec-recall-poisson, and for more details on these representations of particular examples, we refer to the abundant literature on the subject (see [@bremaud1981point, @Cin11] and the references therein).

In the following we consider an individual-based stochastic population $(Z_t)_{t\in [0,T]}$, keeping the notations introduced in @sec-events and @sec-event-intensity for the events and their intensities.
In particular, the set of events types that define the population evolution is denoted by $\mathcal{P} \cup \mathcal{E} \cup \mathcal{E}_W \subset E$, with $\mathcal{P}\,$ the set of events types with Poisson intensity verifying @cnj-intensity-poisson, $\mathcal{E}$ the set of events types with individual intensity and no interaction, verifying @cnj-intensity1 and finally $\mathcal{E}_W$ the set of event types with interactions, verifying @cnj-intensity2.

### Non-explosion criterion {.unnumbered}

First, one has to ensure that the number of events occurring in the population will not explode in finite time, leading to an infinite simulation time. @cnj-intensity1 and @cnj-intensity2 are not sufficient to guarantee the non explosion of the event number, due to the potential explosion of the population size in the presence of interactions. An example is the case when only birth events occur, with an intensity $\Lambda^b_t(Z_t) = C_b (N_t^a)^2$ (i.e. when $W^b(t, I,J) =C_b$). Then, the number of alive individuals $(N_t^a)_{t\geq 0}$ is a well-known pure birth process of intensity function $g(n) = C_b n^2$ (intensity of moving from state $n$ to $n+1$). This process explodes in finite time, since $g$ does not verify the necessary and sufficient non explosion criterion for pure birth Markov processes: $\sum_{n=1}^\infty \frac{1}{g(n)} = \infty$ (see e.g. Theorem 2.2 in [@BanMel15]). There is thus an explosion in finite time of birth events.

@@

@cnj-non-explosion yields that the global intensity $\Lambda_t^e(\cdot)$ of event $e$ is bounded by a function $g^e$ only depending on the population size:
$$
    \Lambda_t^e (Z) \leq g^e(n) := nf^e(n), \quad \text{with }\sum_{n=1}^{\infty} \frac{1}{g^e(n)} = \infty.
$$
If $e\in \mathcal{P}\,$ has a Poisson intensity, then $\Lambda_t^e(Z) =\mu^e_t$ always verifies the previous equation with $g^e(n) = \bar \mu^e$.

@@

The proof of @lem-non-explosion, detailed in Appendix @sec-prf-lem-non-explosion is more technical and relies on a pathwise comparison result, generalizing those obtained in [@KaaElK20]. An alternative pathwise representation of the population process, inspired by the randomized @alg-randomized is given as well in @prp-randomized.

@@

We now present the main algorithm for simulating the evolution of an IBM over $[0,T]$. The algorithm implemented in `IBMPopSim` allows the exact simulation of @eq-sde-eq-Z, based on an acceptance/reject algorithm for simulating random times called *thinning*. The exact simulation of event times with this acceptance/reject procedure is closely related to the simulations of inhomogeneous Poisson processes by the so-called thinning algorithm, often attributed to @LewShe79. 
The simulation methods for inhomogeneous Poisson processes can be adapted to IBMs, and we introduce in this section a general algorithm extending those by @FouMel04 (see also @FerTra09, @Ben10).

The algorithm is based on exponential "candidate" event times, chosen with a (constant) intensity which must be greater than the global event intensity $\Lambda_t(Z_t)$ (@eq-global-intensity). Starting from time $t$, once a candidate event time $t + \bar T_\ell$ has been proposed, a candidate event type $e$ (birth, death,\...) is chosen with a probability $p^e$ depending on the event intensity bounds $\bar \mu^e$, $\bar \lambda^e$ and $\bar W^e$, as defined in @cnj-intensity1 and @cnj-intensity2. 
An individual $I$ is then drawn from the population. 
Finally, it remains to accept or reject the candidate event with a probability $q^e(t,I,Z_t)$ depending on the true event intensity. 
If the candidate event time is accepted, then the event $e$ occurs at time $t + \bar T_\ell$ to the individual $I$. The main idea of the implemented algorithm can be summarized as follows:

@@

Let $(N^j)_{j \in \mathcal{J}}$ be a (inhomogeneous) multivariate Poisson process indexed by a finite set $\mathcal{J}$, such that $\forall j \in \mathcal{J}$, the intensity $(\lambda_j(t))_{t\in [0,T]}$ of $N_j$ is bounded on $[0,T]$:
$$
    \sup_{t\in[0,T]} \lambda_j(t) \le \bar \lambda_j, \text{ and let }
    \bar \Lambda = \sum_{j \in \mathcal{J}} \bar \lambda_j.
$$ 
Recall that such multivariate counting process can be rewritten as a Poisson random measure $N= \sum_{k\geq 1} \delta_{(T_k,J_k)}$ on ${\mathbb{R}}^+\times \mathcal{J}$ (see e.g. Sec. 2 of Chapter 6 in [@Cin11]), where $T_k$ is $k$th jump time of $\sum_{j\in \mathcal{J}} N^j$ and $J_k$ corresponds to the component of the the vector which jumps. In particular, $N^j_t = N([0,t]\times \{j\})$.

@@

\begin{algorithm}
\caption{Thinning algorithm for multivariate inhomogeneous Poisson processes.}
\begin{algorithmic}
    \State \textbf{Input:} Functions and bounds $(\lambda_j, \bar \lambda_j)$, $\lambda_j: [0,T] \to [0,\bar \lambda_j]$ and $\bar \Lambda = \sum_{j \in \mathcal{J}} \bar \lambda_j$
    \State \textbf{Output:} Points $(T_k, J_k)$ of Poisson measure $N$ on $[0,T] \times \mathcal{J}$
    \State Initialization $T_0 \longleftarrow 0$, $\bar T_0 \longleftarrow 0$
    \While{$T_k < T$}
        \Repeat
            \State increment iterative variable $\ell \longleftarrow \ell+1$
            \State compute next proposed time $\bar T_{\ell} \longleftarrow \bar T_{\ell-1} + S_\ell$  with $S_\ell \sim \mathcal{E}(\bar \Lambda)$ 
            \State draw $\bar J_\ell \sim \mathcal{U}\big\{ \bar \lambda_j / \bar \Lambda, j \in \mathcal{J} \big\}$ i.e. $\mathbb{P}(\bar J_\ell = j) = \bar \lambda_j / \bar \Lambda$
            \State draw $\bar \Theta_\ell \sim \mathcal{U}([0,\bar \lambda_{\bar J_\ell}])$
        \Until{accepted event $\bar \Theta_\ell \le \lambda_{\bar J_\ell}(\bar T_{\ell})$} 
        \State record $(T_k, J_k) \longleftarrow (\bar T_{\ell}, \bar J_\ell)$ as accepted point
    \EndWhile
\end{algorithmic}
\end{algorithm}
```

@@

### First event simulation {.unnumbered}

@@

Since the population composition is constant before the first event time, each counting process $N^{j}$ with $j=(k,e) \in \mathcal{J}_0$ coincides on $[0,T_1[$ with an inhomogeneous Poisson process, of intensity $\lambda_t^e (I_k,Z_0)$. Thus (conditionally to $Z_0$), $T_1$ is also the first jump time of an inhomogeneous multivariate Poisson process $N^0 = (N^{0,j})_{j \in \mathcal{J}_0}$ of intensity function $(\lambda_j)_{j\in \mathcal J_0}$, defined for all $j = (k,e) \in \mathcal{J}_0$ by:
$$
    \lambda_j(t) = \lambda^e_t(I_k,Z_0) \le \bar \lambda^e_0 \quad \text{with} \quad \bar \lambda^e_0 = \bar \lambda^e \mathbf{1}_{e \in \mathcal{E}} + \bar W^e N_0 \mathbf{1}_{e \in \mathcal{E}_W},
$$
by @cnj-intensity1 and @cnj-intensity2. In particular, the jump times of $N^0$ occur at the intensity
$$
    \Lambda(t) =\sum_{j \in \mathcal{J}_0} \lambda_j(t)  =\sum_{e \in \mathcal{E} \cup \mathcal{E}_W} \sum_{k=1}^{N_0}  \lambda^e_t(I_k,Z_0) \leq \bar \Lambda(N_0)=N_0 \sum_{e \in \mathcal{E} \cup \mathcal{E}_W} \bar \lambda^e_0.
$$
By @prp-thinning-vector, $N^0$ can be obtained by thinning of the marked Poisson measure $\bar Q^0 = \sum_{\ell \geq 1} \delta_{(\bar T_\ell , (\bar{K}_\ell, \bar E_\ell), \bar \Theta_\ell)}$ on ${\mathbb{R}}^+\times\mathcal{J}_0 \times {\mathbb{R}}^+$, with:

@@

::: {.proof}
The only difference between @alg-interaction and @alg-randomized is in the acceptance/rejection step of proposed events, in the presence of interactions.
In @alg-randomized , a proposed event $(\bar T_\ell, \bar E_\ell, \bar K_\ell)$, with $\bar E_l \in \mathcal{E}_W$ (an event with interaction), is accepted as a true event in the population if
$$
    \bar \Theta_\ell \le W^{\bar E_\ell}(\bar T_\ell, I_{\bar K_\ell}, I_{\bar J_\ell}), \text{ with } (\bar \Theta_\ell, \bar J_\ell) \sim  \mathcal{U}\big([0, \bar W^{\bar E_\ell}] \times \{1, \dots, N_{T_k}\} \big).
$$
By @eq-randomized-event, the probability of accepting this event is the same than in @alg-interaction, which achieves the proof.
:::

@@

The most important step of the model creation is the events creation. The call to the function creating an event is of form
```{R}
#| eval: false 
mk_event_CLASS(type="TYPE", name="NAME", ...)
```
where `CLASS` is replaced by the class of the event intensity, described in @sec-event-intensity , and `type` corresponds to the event type, described in @sec-events.
@tbl-intensity-classes and @tbl-event-types summarize the different possible choices for intensity classes and types of event. The optional argument `name` gives a name to the event. If not specified, the name of the event is its type, for instance `death`. However, a name must be specified if the model is composed of several events with the same type (for instance when there are multiple death events corresponding to different causes of death). The other arguments depend on the intensity class and on the event type.

@@

The proof of @thm-EqZrandomized  follows the same steps than the proof of
@thm-eq-Z.
